Equalizer Programming Guide

### 2007-08-26 Developer/ProgrammingGuide/paper.tex
Update Programming Guide, fix minor issue in website footer, add link to previous polls.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\title{Equalizer Programming Guide}
\author{Stefan Eilemann\thanks{eile@eyescale.ch}\\[\medskipamount]
  % Eyescale Software Sarl
}
\date{
  \textbf{INCOMPLETE}\\[\medskipamount]
  Version 0.2, \today
}

@@

\thispagestyle{empty}
\begin{figure}[ht]
  \centering
  \includegraphics[width=4cm]{logo.pdf}
%  {\small\htmladdnormallink{www.eyescale.ch}{http://www.eyescale.ch}}
\end{figure}
\vfill

@@

\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Version & \bf Date     & \bf Changes \\
    \hline
    0.2         & Aug 20, 2007 & main function described\\
    0.1         & Aug 19, 2007 & outlined the basic concepts\\
    \hline \multicolumn{3}{c}{\small
      \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
      {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}\\
  \end{tabularx}}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\columnwidth]{desk.pdf}{\caption{\label{fConfig}An
      example configuration}}
\end{figure}

\fig{fConfig} shows one example configuration, running on a single
machine (node) using one graphics card (pipe) with one window to render
to two output channels connected to a L-shaped projection table. The
compound description is only used by the server to compute the rendering
tasks. The application is not aware of compounds, and does not need to
concern itself with the parallel rendering logics of a configuration.

@@

The second step is to initialize the Equalizer library. The
initialization function of Equalizer also parses the command line, which
is used to set certain default values based on Equalizer-specific
options\footnote{Equalizer-specific options always start with --eq-},
e.g., the default server location. Furthermore, a node factory is
provided:

@@

\begin{enumerate}
  \item \textbf{Application}: The executable started by the user, which
    is the controlling entity in the rendering session.
  \item \textbf{Auto-launched Render Client}: The typical render client,
    started by the server. In this mode, \textsf{initLocal} never
    returns and the process is terminated by the server.
  \item \textbf{Resident Render Client}: Manually pre-started render
    client, listening on a specified port for server commands.
\end{enumerate}

>>>>

\title{Equalizer Programming Guide}
\author{Stefan Eilemann\thanks{eile@eyescale.ch}\\[\medskipamount]
  % Eyescale Software Sarl
}
\date{
  \textbf{INCOMPLETE}\\[\medskipamount]
  Version 0.3, \today
}

@@

\thispagestyle{empty}
\begin{figure}[ht]
  \hfill
  \includegraphics[width=4cm]{logo.pdf}\hfill
  \includegraphics[width=4cm]{vmml.pdf}\hfill\vspace{-1em}\\
%  {\small\htmladdnormallink{www.eyescale.ch}{http://www.eyescale.ch}}
\end{figure}
\vfill

@@

\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Version & \bf Date     & \bf Changes \\
    \hline
    0.3         & Aug 26, 2007 & application and render client code described\\
    0.2         & Aug 20, 2007 & main function described\\
    0.1         & Aug 19, 2007 & outlined the basic concepts\\
    \hline \multicolumn{3}{c}{\small
      \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
      {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}\\
  \end{tabularx}}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\columnwidth]{cave.pdf}{\caption{\label{fConfig}An
      example configuration}}
\end{figure}

\fig{fConfig} shows one example configuration for a four-side
CAVE\texttrademark, running on two machines (node) using three graphics
cards (pipe) with one window each to render to the four output channels
connected to the projectors for each of the walls. The compound
description is only used by the server to compute the rendering
tasks. The application is not aware of compounds, and does not need to
concern itself with the parallel rendering logics of a configuration.

@@

The second step is to initialize the Equalizer library. The
initialization function of Equalizer also parses the command line, which
is used to set certain default values based on Equalizer-specific
options\footnote{Equalizer-specific options always start with --eq-},
e.g., the default server location. Furthermore, a node factory is
provided. The \textsf{EQERROR} macro, and its counterparts
\textsf{EQWARN}, \textsf{EQINFO} and \textsf{EQVERB} allow selective
debugging outputs with various logging levels:

@@

\begin{enumerate}
  \item \textbf{Application}: The executable started by the user, which
    is the controlling entity in the rendering session.
  \item \textbf{Auto-launched render client}: The typical render client,
    started by the server. The server starts the executable with special
    parameters, which cause \textsf{Client::initLocal} to never
    return. During exit, the server terminates the process.
  \item \textbf{Resident render client}: Manually pre-started render
    client, listening on a specified port for server commands. This mode
    is selected using the command-line option \textsf{--eq-client} and
    potentially \textsf{--eq-listen} and \textsf{-r}\footnote{see \htmladdnormallink{http://www.equalizergraphics.com/documents/design/residentNodes.html}{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
\end{enumerate}

\subsubsection{Main Loop}

The application's main loop starts by connecting the application to an
Equalizer server. The command line parameter \textsf{--eq-server}
explicitly specifies a server address. If no server was specified,
\textsf{Client::connectServer} try first to connect to a server on the
local machine using the default port 4242. If that fails, it will create
a server running within the applications process with a default
1-channel configuration\footnote{see
  \htmladdnormallink{http://www.equalizergraphics.com/documents/design/standalone.html}{http://www.equalizergraphics.com/documents/design/standalone.html}}:

{\small\begin{lstlisting}
int Application::run()
{
    // 1. connect to server
    RefPtr<eq::Server> server = new eq::Server;
    if( !connectServer( server ))
    {
        EQERROR << "Can't open server" << endl;
        return EXIT_FAILURE;
    }
\end{lstlisting}}

The second step is to ask the server for a configuration. The
\textsf{ConfigParams} are a placeholder for later implementations to
provide additional hints and information to the server for choosing a
configuration. The configuration is created using
\textsf{NodeFactory::createConfig}. Therefore it is of type
\textsf{eqPly::Config}, but the return value is \textsf{eq::Config},
making q cast necessary:

{\small\begin{lstlisting}
    // 2. choose config
    eq::ConfigParams configParams;
    Config* config = static_cast<Config*>(server->chooseConfig( configParams ));

    if( !config )
    {
        EQERROR << "No matching config on server" << endl;
        disconnectServer( server );
        return EXIT_FAILURE;
    }
\end{lstlisting}}

Finally it is time to initialize the configuration. For statistics, the
time for this operation is measures and printed. During initialization
the server launches and connects all render client nodes, and calls the
appropriate initialization task methods, as explained in later
sections. \textsf{Config::init} does return after all nodes, pipes,
windows and channels are initialized. It returns \textsf{true} only if
all init task methods were successful. The \textsf{EQLOG} macro allows
topic-specific logging. The numeric topic values are specified in the
respective \textsf{log.h} header files:

{\small\begin{lstlisting}
    // 3. init config
    eqBase::Clock clock;

    config->setInitData( _initData );
    if( !config->init( ))
    {
        EQERROR << "Config initialization failed: " 
                << config->getErrorMessage() << endl;
        server->releaseConfig( config );
        disconnectServer( server );
        return EXIT_FAILURE;
    }

    EQLOG( eq::LOG_CUSTOM ) << "Config init took " << clock.getTimef() << " ms"
                            << endl;
\end{lstlisting}}

When the configuration was successfully initialized, the actual main
loop is executed. The main loop runs until the user exits the
configuration or a maximum number of frames, specified as a command-line
argument, has been rendered. The latter is useful for benchmarks. The
\textsf{Clock} is reused for measuring the overall performance. A new
frame is started using \textsf{Config::startFrame} and a frame is
finished using \textsf{Config::endFrame}.

When the frame is started, the server computes all rendering tasks and
sends them to the appropriate render client nodes. The render client
nodes dispatch the tasks to the correct node or pipe thread, were they
are executed in the order they arrive.

The frame finished is \textsf{latency} frames behind the frame
started. The latency is specified in the configuration file, and allows
several outstanding frames whose tasks are already queued in the node
and pipe threads for execution. This allows overlapped execution and
minimizes idle times.

When the main loop is finished, \textsf{Config::finishAllFrames} catches
up with the latency. It returns after all outstanding frames have been
rendered:

{\small\begin{lstlisting}
    // 4. run main loop
    uint32_t maxFrames = _initData.getMaxFrames();

    clock.reset();
    while( config->isRunning( ) && maxFrames-- )
    {
        config->startFrame();
        // config->renderData(...);
        config->finishFrame();
    }
    const uint32_t frame = config->finishAllFrames();
    const float    time  = clock.getTimef();
    EQLOG( eq::LOG_CUSTOM ) << "Rendering took " << time << " ms (" << frame
                            << " frames @ " << ( frame / time * 1000.f)
                            << " FPS)" << endl;
\end{lstlisting}}

The remainder of the application code cleans up in the reverse order of
the initialization. The config is exited, released and the connection to
the server is closed:

{\small\begin{lstlisting}
    // 5. exit config
    clock.reset();
    config->exit();
    EQLOG( eq::LOG_CUSTOM ) << "Exit took " << clock.getTimef() << " ms" <<endl;

    // 6. cleanup and exit
    server->releaseConfig( config );
    if( !disconnectServer( server ))
        EQERROR << "Client::disconnectServer failed" << endl;
    server = 0;
    return EXIT_SUCCESS;
}
\end{lstlisting}}

\subsubsection{Render Clients}

In the second and third case, when the executable is used as a render
client, \textsf{Client::initLocal} never returns. In order to keep the
client resident, the \textsf{eqPly} example overrides the client loop to
keep it running beyond one configuration run:

{\small\begin{lstlisting}
bool Application::clientLoop()
{
    if( !_initData.isResident( )) // execute only one config run
        return eq::Client::clientLoop();

    // else execute client loops 'forever'
    while( true ) // TODO: implement SIGHUP handler to exit?
    {
        if( !eq::Client::clientLoop( ))
            return false;
        EQINFO << "One configuration run successfully executed" << endl;
    }
    return true;
}
\end{lstlisting}}

\if 0
{\small\begin{lstlisting}
\end{lstlisting}}
\fi
